A Partitioning Algorithm for Maximum Common Subgraph Problems

### 2017-02-17 paper/paper.tex
Less RAM
Expand
Wee edits

We initially assume that graphs are unlabelled, undirected and without loops
(\cref{sec:extensions} describes how these restrictions may be relaxed).
The vertices and edge sets of a graph $\graphG$ are denoted $\V(\graphG)$ and $\E(\graphG)$.  The
set of vertices adjacent to vertex $v$ in graph $\graphG$ is called the
\emph{neighbourhood} of $v$, denoted $\N(\graphG, v)$. We denote by $\invN(\graphG, v)$ the
\emph{inverse neighbourhood} of $v$, being the set of the vertices not adjacent
to $v$ (excluding $v$ itself). A \emph{subgraph} of a graph $\graphG$ is a graph
consisting of some of the vertices of $\graphG$, and all of the edges between these
vertices. A \emph{common subgraph} of two graphs is a graph which is
(isomorphic to) a subgraph of two graphs simultaneously, and a \emph{maximum}
common subgraph is one with as many vertices as possible. (In this paper, all
subgraphs are induced. The \emph{maximum common partial subgraph} problem
instead asks for a common non-induced subgraph with as many \emph{edges} as
possible \citep{DBLP:conf/cp/NdiayeS11}.)

@@

With these definitions established, we now present \McSplit. This algorithm
finds a maximum-cardinality mapping $M^* = \{(v_1, w_1), \dots, (v_{m},
w_{m})\}$ with $|M^*| = m$ vertex pairs, where the $v_i$ are distinct members
of $\V(\graphG)$ and the $w_i$ are distinct vertices of $\V(\graphH)$, such
that $v_i$ and $v_j$ are adjacent in $\graphG$ if and only if $w_i$ and $w_j$
are adjacent in $\graphH$.  Given such a mapping, the subgraph of $\graphG$
induced by $\{v_i, \dots, v_{m}\}$ and the subgraph of $\graphH$ induced by
$\{w_i, \dots, w_{m}\}$ are a maximum common subgraph.

\paragraph{Walkthrough} Before discussing the algorithm in detail, we start
with an illustration of the main concepts using the graphs $\graphG$ and $\graphH$ in
\cref{fig:alg1}.  These graphs have a maximum common subgraph with four
vertices; one example is the mapping $\{1a, 2f, 3d, 5b\}$ where vertex $1$ is
assigned to vertex $a$, $2$ is assigned to $f$, $3$ to $d$ and $5$ to $b$.

@@

We begin by mapping vertex $1$ to vertex $a$, giving $M=\{1a\}$.  Now label
each unmatched vertex according to whether it is adjacent to vertex $1$ (in
$\graphG$) or vertex $a$ (in $\graphH$), as shown in \cref{fig:alg2}.  Adjacent
vertices have label 1; non-adjacent vertices have label 0.  We can extend $M$
with a vertex mapping $vw$, with $v \in \V(\graphG)$ and $w \in \V(\graphH)$, if and only
if $v$ and $w$ have the same label.  This property, that two vertices may be
mapped together if and only if they share a label, is the algorithm's main
invariant.

@@

\begin{algorithm}[t]
\DontPrintSemicolon
\nl $\FuncSty{search}(\AlgVar{future},M)$ \;
\nl \Begin{
%\nl \lIf {$\AlgVar{future} = \emptyset$ \bf{and} $|M| > |\AlgVar{incumbent}|$}
\nl \lIf {$|M| > |\AlgVar{incumbent}|$}{$\AlgVar{incumbent} \gets M$} \label{StoreIncumbent}
%\nl \lIf {$\AlgVar{future} = \emptyset$}{return}
\medskip
\nl $\AlgVar{bound} \gets |M|  + \sum_{\langle \setG,\setH \rangle \in \AlgVar{future}} \min(|\setG|,|\setH|)$ \label{CalcBound} \;
\nl \lIf {$\AlgVar{bound} \leq |\AlgVar{incumbent}|$}{\KwSty{return}} \label{PruneSearch}
\medskip
\nl $\langle \setG,\setH \rangle \gets \FuncSty{SelectLabelClass}(\AlgVar{future})$ \label{SelectClass} \;
\nl $v \gets \FuncSty{SelectVertex}(\setG)$ \label{SelectVertex} \;
\nl \For {$w \in \setH$ \label{WLoop}} {
    \nl    $M \gets M \cup \{(v,w)\}$ \label{GrowM} \;
\nl    \bf{Let} $\AlgVar{future'} \gets \emptyset$ \label{NewFuture} \;
\nl    \For {$\langle \setG',\setH'\rangle \in future$ \label{InnerLoop}}{
\nl        \bf{Let} $\setG'' \gets \setG' \cap \N(\graphG, v) \setminus \{v\}$ \label{NewPWithEdge} \;
\nl        \bf{Let} $\setH'' \gets \setH' \cap \N(\graphH, w) \setminus \{w\}$ \;
\nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} + \langle \setG'' , \setH'' \rangle$ \label{AddToFutureWithEdge}}
\nl        \bf{Let} $\setG'' \gets \setG' \cap \invN(\graphG, v) \setminus \{v\}$ \label{NewPWithoutEdge}  \;
\nl        \bf{Let} $\setH'' \gets \setH' \cap \invN(\graphH, w) \setminus \{w\}$ \;
\nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} + \langle \setG'' , \setH'' \rangle$} \label{InnerLoopEnd}
       }
\nl   $\FuncSty{search}(\AlgVar{future'},M)$ \label{ExpandWithV} \;
\nl   $M \gets M \setminus \{(v,w)\}$ \label{ShrinkM} \;
  }
\nl $\setG' \gets \setG \setminus \{v\}$ \label{RemoveV} \;
\nl $\AlgVar{future} \gets \AlgVar{future} \setminus \{\langle \setG,\setH \rangle\}$\;
\nl \lIf {$\setG' \neq \emptyset$} {$\AlgVar{future} \gets \AlgVar{future} \cup \{\langle \setG',\setH \rangle \}$}
\nl $\FuncSty{search}(\AlgVar{future},M)$ \label{ExpandWithoutV} \;
}
\;
\nl $\FuncSty{McSplit}(\graphG,\graphH)$ \label{McSplitFun} \;
\nl \Begin{
    \nl $\KwSty{global}~\AlgVar{incumbent} \gets \emptyset$ \;
\nl $\FuncSty{search}(\{\langle V(\graphG),V(\graphH) \rangle \},\emptyset)$ \label{FirstExpandCall} \;
\nl $\KwSty{return}$~$\AlgVar{incumbent}$ \;
}
\caption{Finding a maximum common subgraph.}
\label{McSplitAlg}
\end{algorithm}

\paragraph{\cref{McSplitAlg} in detail} The recursive procedure,
$\FuncSty{search}$, has two parameters.  The parameter $\AlgVar{future}$ is a
list of label classes, each represented as a $\langle \setG, \setH \rangle$ pair as
described above.  The parameter $M$ is the current mapping of vertices.  On
each call to $\FuncSty{search}$, the invariant holds that a $(v,w)$ pair may be
added to $M$ if and only if $v$ and $w$ belong to the same label class in
$\AlgVar{future}$.

@@

The remainder of $\FuncSty{search}$ performs the search.  A label class
$\langle \setG, \setH \rangle$ is selected from $\AlgVar{future}$, where $\setG \subseteq
V(\graphG)$ and $\setH \subseteq V(\graphH)$, using some heuristic (\lineref{SelectClass}).
From this label class, a vertex $v$ is selected from $\setG$ using a heuristic
(\lineref{SelectVertex}). We now iterate over all possible vertices in $\setH$
(\linerangeref{WLoop}{ShrinkM}). A vertex $w \in \setH$ is selected and mapping $M$
is extended (\lineref{GrowM}). A new set of label-classes, $\AlgVar{future'}$,
is created (\lineref{NewFuture}).  Every label-class in $\AlgVar{future}$ can
now be split (\linerangeref{InnerLoop}{InnerLoopEnd}) into two new classes. The
first of these classes (\linerangeref{NewPWithEdge}{AddToFutureWithEdge})
contains vertices in $\setG$ adjacent to $v$ and vertices in $\setH$ adjacent to $w$.
This is added to $\AlgVar{future'}$ if both sets contain at least one vertex.
This is then repeated symmetrically for non-adjacency
(\linerangeref{NewPWithoutEdge}{InnerLoopEnd}). A recursive call is then made
(\lineref{ExpandWithV}), on return from which we remove the mapping $(v,w)$.
Having explored all possible mappings of $v$ with vertices in $\setH$ we now
consider what happens if $v$ is not matched
(\linerangeref{RemoveV}{ExpandWithoutV}).

We start our search at the function $\FuncSty{McSplit}$ (\lineref{McSplitFun}),
with graphs $\graphG$ and $\graphH$ as inputs.  This function returns the size of the
largest mapping.  In \lineref{FirstExpandCall} the initial call is made to
$\FuncSty{search}$; at this point we have a single label-class containing all
vertices, and the mapping $M$ is empty.

@@

Experiments were performed on machines with dual Intel Xeon E5-2640 v2 CPUs and
64GBytes RAM. Our algorithm was implemented\footnote{URL removed for anonymous
review} in C++ and compiled using g++ 5.3.0. We compare against
the best constraint programming implementations of
\citet{DBLP:conf/cp/NdiayeS11} and \citet{DBLP:conf/cp/McCreeshNPS16} (CP-FC in
the unlabelled cases, and CP-MAC in the labelled cases, using both branching
and filtering for connected subgraphs), the clique encodings of
\citet{DBLP:conf/cp/McCreeshNPS16}, and the $k{\downarrow}$ algorithm of
\citet{UpcomingAAAIPaper} (which only supports unlabelled, undirected,
unconnected instances).

@@

\paragraph{Unlabelled, undirected, connected} This class of instances are shown
in \cref{figure:plain-connected-cumulative}.  These results are very similar to
the corresponding experiment in \cref{figure:plain-cumulative} in which the
subgraph is not required to be connected: \McSplit\ is the clear winner by more
than an order of magnitude.

>>>>

We initially assume that graphs are unlabelled, undirected and without loops
(\cref{sec:extensions} describes how these restrictions may be relaxed).
The vertex and edge sets of a graph $\graphG$ are denoted $\V(\graphG)$ and $\E(\graphG)$.  The
set of vertices adjacent to vertex $v$ in graph $\graphG$ is called the
\emph{neighbourhood} of $v$, denoted $\N(\graphG, v)$. We denote by $\invN(\graphG, v)$ the
\emph{inverse neighbourhood} of $v$, being the set of the vertices not adjacent
to $v$ (excluding $v$ itself). A \emph{subgraph} of a graph $\graphG$ is a graph
consisting of some of the vertices of $\graphG$, and all of the edges between these
vertices. A \emph{common subgraph} of two graphs is a graph which is
(isomorphic to) a subgraph of two graphs simultaneously, and a \emph{maximum}
common subgraph is one with as many vertices as possible. (In this paper, all
subgraphs are induced. The \emph{maximum common partial subgraph} problem
instead asks for a common non-induced subgraph with as many \emph{edges} as
possible.)

@@

With these definitions established, we now present \McSplit. This algorithm
finds a maximum-cardinality mapping $M^* = \{(v_1, w_1), \dots, (v_{m},
w_{m})\}$ with $|M^*| = m$ vertex pairs, where the $v_i$ are distinct vertices
from $\V(\graphG)$ and the $w_i$ are distinct vertices from $\V(\graphH)$, such
that $v_i$ and $v_j$ are adjacent in $\graphG$ if and only if $w_i$ and $w_j$
are adjacent in $\graphH$.  Given such a mapping, the subgraph of $\graphG$
induced by $\{v_i, \dots, v_{m}\}$ and the subgraph of $\graphH$ induced by
$\{w_i, \dots, w_{m}\}$ are a maximum common subgraph.

\paragraph{Walkthrough} Before discussing the algorithm in detail, we illustrate
the main concepts using the graphs $\graphG$ and $\graphH$ in
\cref{fig:alg1}.  These graphs have a maximum common subgraph with four
vertices; one example is the mapping $\{1a, 2f, 3d, 5b\}$ where vertex $1$ is
assigned to vertex $a$, $2$ is assigned to $f$, $3$ to $d$ and $5$ to $b$.

@@

We begin by mapping vertex $1$ to vertex $a$, giving $M=\{1a\}$.  Now
label each unmatched vertex in $\V(\graphG)$ according to whether it is adjacent to vertex $1$, and
label each unmatched vertex in $\V(\graphH)$ according to whether it is adjacent to vertex $a$,
as shown in \cref{fig:alg2}.  Adjacent
vertices have label 1; non-adjacent vertices have label 0.  We can extend $M$
with a mapping $vw$, with $v \in \V(\graphG)$ and $w \in \V(\graphH)$, if and only
if $v$ and $w$ have the same label.  This property, that two vertices may be
mapped together if and only if they share a label, is the algorithm's main
invariant.

@@

\begin{algorithm}[t]
\DontPrintSemicolon
\nl $\FuncSty{expand}(\AlgVar{future},M)$ \;
\nl \Begin{
%\nl \lIf {$\AlgVar{future} = \emptyset$ \bf{and} $|M| > |\AlgVar{incumbent}|$}
\nl \lIf {$|M| > |\AlgVar{incumbent}|$}{$\AlgVar{incumbent} \gets \AlgVar{M}$} \label{StoreIncumbent}
%\nl \lIf {$\AlgVar{future} = \emptyset$}{return}
\medskip
\nl $\AlgVar{bound} \gets |M|  + \sum_{\langle \setG,\setH \rangle \in \AlgVar{future}} \min(|\setG|,|\setH|)$ \label{CalcBound} \;
\nl \lIf {$\AlgVar{bound} \leq |\AlgVar{incumbent}|$}{return} \label{PruneSearch}
\medskip
\nl $\langle \setG,\setH \rangle \gets \FuncSty{SelectLabelClass}(\AlgVar{future})$ \label{SelectClass} \;
\nl $v \gets \FuncSty{SelectVertex}(\setG)$ \label{SelectVertex} \;
\nl \For {$w \in \setH$ \label{WLoop}} {
    \nl    $M \gets M \cup \{(v,w)\}$ \label{GrowM} \;
\nl    \bf{Let} $\AlgVar{future'} \gets \emptyset$ \label{NewFuture} \;
\nl    \For {$\langle \setG',\setH'\rangle \in future$ \label{InnerLoop}}{
\nl        \bf{Let} $\setG'' \gets \setG' \cap \N(\graphG, v) \setminus \{v\}$ \label{NewPWithEdge} \;
\nl        \bf{Let} $\setH'' \gets \setH' \cap \N(\graphH, w) \setminus \{w\}$ \;
\nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} \cup \{\langle \setG'' , \setH'' \rangle\}$ \label{AddToFutureWithEdge}}
\nl        \bf{Let} $\setG'' \gets \setG' \cap \invN(\graphG, v) \setminus \{v\}$ \label{NewPWithoutEdge}  \;
\nl        \bf{Let} $\setH'' \gets \setH' \cap \invN(\graphH, w) \setminus \{w\}$ \;
\nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} \cup \{\langle \setG'' , \setH'' \rangle\}$} \label{InnerLoopEnd}
       }
\nl   $\FuncSty{expand}(\AlgVar{future'},M)$ \label{ExpandWithV} \;
\nl   $M \gets M \setminus \{(v,w)\}$ \label{ShrinkM} \;
  }
\nl $\setG' \gets \setG \setminus \{v\}$ \label{RemoveV} \;
\nl $\AlgVar{future} \gets \AlgVar{future} \setminus \{\langle \setG,\setH \rangle\}$\;
\nl \lIf {$\setG' \neq \emptyset$} {$\AlgVar{future} \gets \AlgVar{future} \cup \{\langle \setG',\setH \rangle \}$}
\nl $\FuncSty{expand}(\AlgVar{future},M)$ \label{ExpandWithoutV} \;
}
\;
\nl $\FuncSty{McSplit}(\graphG,\graphH)$ \label{McSplitFun} \;
\nl \Begin{
\nl $\AlgVar{incumbent} \gets \emptyset$ \;
\nl $\FuncSty{expand}(\{\langle V(\graphG),V(\graphH) \rangle \},\emptyset)$ \label{FirstExpandCall} \;
\nl return $|\AlgVar{incumbent}|$ \;
}
\caption{Finding a maximum common subgraph.}
\label{McSplitAlg}
\end{algorithm}

\paragraph{\cref{McSplitAlg} in detail} The recursive procedure,
$\FuncSty{expand}$, has two parameters.  The parameter $\AlgVar{future}$ is a
list of label classes, each represented as a $\langle \setG, \setH \rangle$ pair as
described above.  The parameter $M$ is the current mapping of vertices.  On
each call to $\FuncSty{expand}$, the invariant holds that a $(v,w)$ pair may be
added to $M$ if and only if $v$ and $w$ belong to the same label class in
$\AlgVar{future}$.

@@

The remainder of $\FuncSty{expand}$ performs the search.  A label class
$\langle \setG, \setH \rangle$ is selected from $\AlgVar{future}$, where $\setG \subseteq
V(\graphG)$ and $\setH \subseteq V(\graphH)$, using some heuristic (\lineref{SelectClass}).
From this label class, a vertex $v$ is selected from $\setG$ using a heuristic
(\lineref{SelectVertex}). We now iterate over all possible vertices in $\setH$
(\linerangeref{WLoop}{ShrinkM}). A vertex $w \in \setH$ is selected and mapping $M$
is extended (\lineref{GrowM}). A new set of label-classes, $\AlgVar{future'}$,
is created (\lineref{NewFuture}).  Every label-class in $\AlgVar{future}$ can
now be split (\linerangeref{InnerLoop}{InnerLoopEnd}) into two new classes. The
first of these classes (\linerangeref{NewPWithEdge}{AddToFutureWithEdge})
contains vertices in $\setG$ adjacent to $v$ and vertices in $\setH$ adjacent to $w$.
This is added to $\AlgVar{future'}$ if both sets contain at least one vertex.
This is then repeated symmetrically for non-adjacency
(\linerangeref{NewPWithoutEdge}{InnerLoopEnd}). A recursive call is then made
(\lineref{ExpandWithV}), on return from which we remove the mapping $(v,w)$.
Having explored all possible mappings of $v$ with vertices in $\setH$ we now
consider what happens if $v$ is not matched
(\linerangeref{RemoveV}{ExpandWithoutV}).

We start our search at the function $\FuncSty{McSplit}$ (\lineref{McSplitFun}),
with graphs $\graphG$ and $\graphH$ as inputs.  This function returns the size of the
largest mapping.  In \lineref{FirstExpandCall} the initial call is made to
$\FuncSty{expand}$; at this point we have a single label-class containing all
vertices, and the mapping $M$ is empty.

@@

Experiments were performed on machines with dual Intel Xeon E5-2640 v2 CPUs and
64GBytes RAM. Our algorithm was implemented\footnote{URL removed for anonymous
review} in C++ and compiled using g++ 5.3.0. For comparison, we compare against
the best constraint programming implementations of
\citet{DBLP:conf/cp/NdiayeS11} and \citet{DBLP:conf/cp/McCreeshNPS16} (CP-FC in
the unlabelled cases, and CP-MAC in the labelled cases, using both branching
and filtering for connected subgraphs), the clique encodings of
\citet{DBLP:conf/cp/McCreeshNPS16}, and the $k{\downarrow}$ algorithm of
\citet{UpcomingAAAIPaper} (which only supports unlabelled, undirected,
unconnected instances).

@@

\paragraph{Unlabelled, undirected, connected} This class of instances are shown
in \cref{figure:plain-connected-cumulative}.  These results are very similar to
the corresponding experiment in \cref{figure:plain-cumulative} in which the
subgraph is not required to be connected.
